---
title: "Convenience Functions"
author: "Kelly Stratton, Lisa Bramer"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
    rmarkdown::html_vignette:
    fig_caption: yes
    self_contained: yes
    toc: true
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{pmartR-convenience_functions} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Overview

Certain functions have been included in `pmartR` to simplify the user's experience by allowing them to retrieve information stored in attributes of the omicsData object. 


## Related to Data Structure or Setup

- `get_check_names()` - Retrieves the value in the check.names attribute from an omicsData object

```{r}

```


- `set_check_names()` - Sets the check.names attribute of an omicsData object

```{r}

```


- `get_data_class()` - Returns data_class attribute from statRes or trellData object, inherited from the omicsData used in `imd_anova()` or `format_data()`

```{r}

```


- `get_data_info()` - Returns a list containing the data scale, normalization information, number of unique entries in e_data, number of missing observations in e_data, proportion of missing observations in e_data, number of samples, and data type

```{r}

```



- `get_data_scale()` - Returns current data scale which may be different from the original data scale (if `edata_transform()` was used)

```{r}

```



- `get_data_scale_orig()` - Retrieves the character string indicating the scale the data was originally on when the omicsData object was created

```{r}

```



- `get_edata_cname()` - Returns the name of the column in e_data that contains the biomolecule IDs

```{r}

```



- `get_emeta_cname()` - Returns the name of the column in e_meta that contains the mapping to biomolecules in e_data

```{r}

```



- `get_fdata_cname()` - Returns the name of the column in f_data that contains the names of the samples

```{r}

```

- `get_group_DF()` - A data.frame with columns for sample ID and group. If two main effects are provided the original main effect levels for each sample are returned as the third and fourth columns of the data frame. Additionally, the covariates provided will be listed as attributes of this data frame.


```{r}

```



- `get_group_table()` - This function returns a table with number of samples per group

```{r}

```

- `get_isobaric_info()` - A list containing the following six elements: exp_cname, channel_cname, refpool_channel, refpool_cname, refpool_notation, and norm_info (list containing a single logical element that indicates whether the data have been normalized to a reference pool


```{r}

```



- `get_meta_info()` - Retrieves the values in the meta_info attribute from an omicsData object

```{r}

```

- `get_nmr_info()` - A list containing the following three elements: metabolite_name, sample_property_cname, and norm_info (list containing two logical elements that indicate i) whether the data have been normalized to a spiked in metabolite or to a property taking sample-specific values and ii) whether the data have been back transformed so the values are on a similar scale to the raw values before normalization.


```{r}

```



## Related to Filters

- `get_filters()` - A list containing filter class objects. Each element in this list corresponds to a filter applied to the data, and filters are listed in the order they were applied.

```{r}

```



## Related to Normalization

- `get_data_norm()` - Returns the norm_info element of the data_info attribute indicating whether the data have been normalized

```{r}

```


- `get_isobaric_norm()` - A logical value indicating whether the data have been isobaric normalized

```{r}

```


- `get_nmr_norm()` - A logical value indicating whether the data have been NMR normalized

```{r}

```


## Related to Statistical Comparisons


- `get_comparisons()` - Returns a data frame with comparisons and their indices

```{r}

```














## References
