---
title: "Data Normalization"
author: "Kelly Stratton, Lisa Bramer"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
    rmarkdown::html_vignette:
    fig_caption: yes
    self_contained: yes
    toc: true
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{pmartR-normalization} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(rmarkdown.html_vignette.check_title = FALSE)

library(pmartR)
library(pmartRdata)
```

## Overview

There are two general types of normalization approaches included in `pmartR`:

1. Instrument normalization - applicable to labeled proteomics data and (sometimes) NMR data 

2. Statistical normalization - applicable to all of the data types supported in `pmartR`, with RNAseq data being a special case

When instrument normalization is required, this is done after quality control (filtering) and prior to statistical normalization.


## Instrument Normalization

### Labeled Proteomics

Labeled proteomics data, such as that generated via TMT (tandem mass tag) or iTRAQ (isobaric tags for relative quantification), involves running subsets of samples from an experiment on separate "plexes" or "plates" or "experiments". Each plex includes a "reference pool" sample, consisting of a pool of all of the experimental samples. (Note that the assignment of samples to a plex is a key part of the randomization and experimental design processes, due to the way data goes missing when these labeled methods are utilized.) Samples must be normalized to their corresponding reference pool in order to enable comparisons between samples.

#### Usage

When working with labeled proteomics data, an `isobaricpepData` object should be used. To perform the reference pool normalization, use the `normalize_isobaric()` function on data that has already been log transformed. 

```{r load_example_isobaricpepData}
data(isobaric_object)
class(isobaric_object)

isobaric_object <- edata_transform(isobaric_object, 
                                   data_scale = "log2")
```


There are two ways to specify the information needed for identifying reference samples which should be used for normalization: 

1. Specify the `refpool_cname` and `refpool_notation`. This should be used when the reference sample is not in a consistent channel across experiments/plates. Here, `refpool_cname` gives the name of the column in `f_data` which indicates whether a sample is a reference or not, and `refpool_notation` is a character string giving the value used to denote a reference sample in that column. } In both cases you must specify `exp_cname` which gives the column name for the column in `f_data` containing information about which plex/plate/experiment a sample was run on.


```{r specify_refpool}
# do not apply the normalization until we have looked at the distribution of peptides within the reference pool samples
iso_norm <- normalize_isobaric(isobaric_object, 
                               exp_cname = "Plex",
                               apply_norm = FALSE,
                               refpool_cname = "Virus",
                               refpool_notation = "Pool")

# look at the distribution of peptides within the reference pool samples
plot(iso_norm)

# now apply the normalization
isobaric_object_normalized <- normalize_isobaric(isobaric_object, 
                               exp_cname = "Plex",
                               apply_norm = TRUE,
                               refpool_cname = "Virus",
                               refpool_notation = "Pool")

# look at boxplots of the data after normalization to the reference pool samples
plot(isobaric_object_normalized)
```

2. Specify the `channel_cname` and `refpool_channel`. This should be used when the reference sample for each plex/plate/experiment was always located in the same channel. Here `channel_cname` gives the column name for the column in `f_data` which gives information about which channel each sample was run on, and `refpool_channel` is a character string specifying the value in `channel_colname` that corresponds to the reference sample channel. 

```{r specify_channels, eval = FALSE}
# not run because this data does not actually contain channel_cname or refpool_channel information
iso_norm <- normalize_isobaric(isobaric_object, 
                               exp_cname = "Plex",
                               apply_norm = FALSE,
                               channel_cname = "SampleChannel", # this column in f_data would specify a number corresponding to the channel for each sample
                               refpool_channel = 4) # this value in the SampleChannel column would always correspond to a reference pool sample
```


### NMR

In order to ensure that samples are comparable to one another, normalizing NMR data to either a sample-specific property (such as sample mass) or to a spiked-in reference standard.

#### Usage

There are two ways to specify the information needed for performing instrument normalization on an nmrData object: 

1. Specify `sample_property_cname`. This should be used when normalizing to a sample property, such as sample concentration or mass, is desired. Here, `sample_property_cname` gives the name of the column in `f_data` which contains the property to use for normalization. If any samples have a missing value for this column, and error is returned. 

2. Specify `metabolite_name`. This should be used when normalization to a spiked in standard is desired. Here `metabolite_name` gives the name of the metabolite in e_data (and e_meta, if present) corresponding to the spiked in standard. If any samples have a missing value for this metabolite, an error is returned. 

The option to backtransform the data after either of these NMR normalizations is offered, if the user would like to ensure values are on a scale similar to their raw values before normalization. The following values are calculated and/or applied for backtransformation purposes: 

- If normalization using a metabolite in `e_data` is specified, the location parameter is the median of the values for  `metabolite_name` 

- If normalization using a sample property in `f_data` is specified, the location parameter is the median of the values in `sample_property`

We'll use the example NMR dataset from the `pmartRdata` package.

```{r}
data(nmr_object_identified)
summary(nmr_object_identified)

# log2 transform the values
nmr_object <- edata_transform(nmr_object_identified, "log2")
```


**Normalize using property of each sample**

We can first create a normRes object and explore the values of the sample property to be used for instrument normalization. The figure below shows the Concentration values for each sample.

```{r}
# don't apply the normalization yet
normRes_property <-
  normalize_nmr(nmr_object,
                apply_norm = FALSE,
                sample_property_cname = "Concentration")
plot(normRes_property)
```

If we are satisified with using these Concentration values for normalization, we can apply the instrument normalization and plot the resulting omicsData object.

```{r }
# now apply the normalization
nmr_norm_property <-
  normalize_nmr(nmr_object,
                apply_norm = TRUE,
                sample_property_cname = "Concentration",
                backtransform = TRUE)
plot(nmr_norm_property)
```

**Normalize using reference metabolite**

We can first create a normRes object and explore the values of the reference metabolite to be used for instrument normalization. The figure below shows the log2 values of the reference metabolite for each sample.

```{r}
# don't apply the normalization yet
normRes_reference <-
  normalize_nmr(nmr_object,
                apply_norm = FALSE,
                metabolite_name = "unkm1.53")
plot(normRes_reference)
```

If we are satisified with using these values for normalization, we can apply the instrument normalization and plot the resulting omicsData object.

```{r}
# now apply the normalization
nmr_norm_reference <- normalize_nmr(nmr_object,
                                    apply_norm = TRUE,
                                    metabolite_name = "unkm1.53")
plot(nmr_norm_reference)
```



## Statistical Normalization

### Global Normalization

There are two specifications needed in order to compute normalization factors, the normalization function (how the normalization factors are computed) and the subset function (the subset of biomolecules with which the normalization factors are computed). Global normalizations in `pmartR` normalize within each sample, so each sample has its own normalization factor(s), a location and (sometimes) a scale parameter.  There are several options available for normalization and subset functions.

**Normalization Functions**

For all of the normalization functions, `na.rm = TRUE` is utilized when computing the parameters.

- Median: Location parameter for each sample is equal to the median of the observed biomolecules in the subset specified. There is no scale parameter.

- Mean: Location parameter for each sample is equal to the mean of the observed biomolecules in the subset specified. There is no scale parameter.

- Median Absolute Deviation (MAD): Location parameter for each sample is equal to the median of the observed biomolecules in the subset specified. Scale parameter is equal to the MAD of the observed biomolecules in the subset specified.

- Z-score: Location parameter for each sample is equal to the mean of the observed biomolecules in the subset specified. Scale parameter is equal to the standard deviation of the observed biomolecules in the subset specified.

**Subset Functions**

- All: All biomolecules are utilized

- L-Order Statistics (LOS): Identifies the subset of the biomolecules associated with the top L order statistics, where L is a proportion between 0 and 1. Specifically, the biomolecules with the top L proportion of highest absolute abundance are retained for each sample, and the union of these biomolecules is taken as the subset identified (Wang et al., 2006)

- Percentage of Peptides Present (PPP): Originally developed for use with proteomics data, this subset can also be applied to other omics data types. PPP identifies the subset of biomolecules that are present/non-missing for a minimum proportion of samples (Karpievitch et al., 2009; Kultima et al., 2009).

- Rank Invariant Peptides (RIP): Identifies biomolecules with complete data that have a p-value greater than a defined threshold alpha (common values include 0.1 or 0.25) when subjected to a Kruskal-Wallis test based (non-parametric one-way ANOVA) on group membership (Webb-Robertson et al., 2011).

- PPP-RIP: Equivalent to RIP, however, rather than requiring biomolecules with complete data, biomolecules with at least a specified proportion of non-missing values are subject to the Kruskal-Wallis test. 

- Complete Peptides: Biomolecules with no missing data across all samples, equivalent to PPP with proportion = 1. 

#### Usage

Global normalization methods are applied with the `normalize_global()` function. There is an argument to the function,  `apply_norm`, that dictates whether the normalization method is actually applied to the data (`apply_norm = TRUE)` or if instead the normalization factors are computed so that the user can assess the appropriateness of the normalization approach prior to applying it to a dataset (`apply_norm = FALSE)`. Similar to NMR normalization, the option to backtransform the data after global normalization is available in order to put the normalized values back on roughly the same scale as the raw (un-normalized) data (`backtransform = TRUE`).

Global median centering is our go-to approach, especially for LC-MS lipidomics, GC-MS metabolomics, and NMR. Below we provide a variety of examples of using the `normalize_global()` function with different parameter values.

When, in preparation for normalizing the data, we use the `normalize_global()` function but do not apply the normalization, we get back an object of class normRes.

```{r}
# global median centering - don't apply the norm, show info in the normRes object


```


When we apply the normalization to the data, we get back an omicsData object of the same class that we started with.

```{r}
# global median centering - apply the norm


```



```{r}
# RIP mean centering - don't apply the norm (so we can look at the number of molecules in the subset)


```


```{r}
# LOS MAD


```


```{r}
# PPP z-score


```


```{r}
# PPP-RIP median


```


```{r}
# Complete mean


```


```{r}
# 


```


```{r}
# 


```


For proteomics data, or large lipidomics or metabolomics datasets (>200ish biomolecules??), there is a method to help select an appropriate normalization approach, Statistical Procedure for the Analyses of peptide abundance Normalization Strategies (SPANS) [@webb2011statistical]. This algorithm ranks different combinations of subset and normalization functions based on a score that captures how much bias a particular normalization procedure introduces into the data.  Higher scores imply less bias.


```{r}

```



### LOESS Normalization

As an alternative to a global normalization, `pmartR` includes a wrapper for the normalizeCyclicLoess [@bolstad2003comparison] function from the limma package [@smyth2005limma]. 

```{r}

```


### Quantile Normalization

Quantile normalization is an algorithm for normalizing a set of data vectors by giving them the same distribution [@bolstad2003comparison]. It is applied to data on the abundance scale (e.g. not a log scale). It is often used for microarray data.

```{r}

```



### RNAseq Data Normalization

The statistical normalization approaches for RNAseq data are combined with the statistical analysis functions. The following are supported within `pmartR` via wrapper functions for the original functionality of the corresponding packages.

- DESeq2 [@love2014differential]

- edgeR [@robinson2010edger]

- limma voom [@law2014voom]


#### DESeq2


```{r}

```


#### edgeR


```{r}

```


#### limma voom


```{r}

```


## References
